pdp: An R Package for Constructing Partial Dependence Plots

### 2017-01-04 manuscript/reviews/response.Rmd

```{r, include=FALSE}
rm(data)
```

### Comment 1.5
I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
`"Error in train[[x]] : subscript out of bounds":`
```{r, eval=FALSE}
data(trees)
foo <- function(data) {
  lm(Volume ~ ., data = data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var = "Girth", plot = TRUE)  # fails
```
and the following silently gives misleading results:

```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)  # ok
trees <- trees[1:2, ]
partial(mod, pred.var = "Girth", plot = TRUE)  # wrong xlim
```
I'm not necessarily saying that you need to resolve the above two
issues, but at least mention in the documentation that this kind of
thing can happen, and how to avoid it (basically call partial in the
same environment used to build the model, and don't change any of the
data used to build the model).

### Response 1.5
Agreed! Both the manuscript (see page 3) and the help page for `partial` have been expanded to discuss this. Unfortunately, some functions don't store a copy of the training data. However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary. Re-running your first example should give an informative error message:
```{r, eval=TRUE, error=TRUE}
data(trees)
foo <- function(data)
{
  lm(Volume ~ ., data = data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var = "Girth", plot = TRUE)  # fails with better error mssg
```

@@

### Comment 1.6
The following code gives an obscure error message:
```
"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
   to = max(train[[x]],  :
  'from' cannot be NA, NaN or infinite"
```
Some better handholding for the user may be helpful here.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
```

### Response 1.6
This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
```{r, error=TRUE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
```

@@

### Comment 1.7
On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`". What happens if `y` is one of the predictors?

### Response 1.7
The last column is now labeled `yhat`, which is less likely to be a predictor name. When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not. I think regardless of what the default column name is, there is always the possibility of this problem. This has been filed as an issue on the `pdp` GitHib page: https://github.com/bgreenwell/pdp/issues/22.

@@

### Comment 1.9
On page 3 in the PDF of the paper, add a newline to the error message in the paper:
```
Error: The training data could not be extracted from object. Please
supply the raw training data using the ...
```

### Response 1.9
Fixed. Thanks for pointing that out.

### Comment 1.10
Is there any reason the package is called `pdp` but the main function is called `partial`?  Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
```{r, eval=FALSE}
library(pdp)
?pdp  # fails
```

### Response 1.10
The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letters). So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names. However, I did add a help page describing the package, main functions, etc. So, `?pdp` no longer fails. However, I am not opposed to creating an alias to `partial` called `pdp` in the future.

@@

### Comment 1.11
`partial` should return the same value regardless of whether or not the `plot` argument is used?

### Response 1.11
I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE`; that is, I want it to return a `"trellis"` object. However, I did add the partial dependence data as an attribute to the returned `"trellis"` object so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial). For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
pd.Girth <- partial(mod, pred.var = "Girth")  # data only
pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
```
Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded. This would be useful if the user did not store the PDP but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern). For example,
```{r, fig.width=4, fig.height=4}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```
This has been noted in the package documentation (see `?partial`). Thanks for the idea!

### Comment 1.12
A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
```{r, eval=FALSE}
par(mfrow = c(1, 2))
partial(ozone.rf, pred.var = "Temp", plot = TRUE)
partial(ozone.rf, pred.var = "Wind", plot = TRUE)
```

### Response 1.12
The command `par(mfrow = c(2, 2))` does not work with `"trellis"` objects, such as those produced by `lattice`---which `partial` relies on for producing its plots. Other methods are available. For convenience, `pdp` now imports the `grid.arrange` function from the `gridxtra` package (this has been discussed in the manuscript and incorporated into the examples where appropriate).
```{r, include=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data = airquality, na.action = na.omit)
library(pdp)
```
```{r, eval=TRUE, fig.width=8, fig.height=4}
grid.arrange(
  partial(ozone.rf, pred.var = "Temp", plot = TRUE),
  partial(ozone.rf, pred.var = "Wind", plot = TRUE),
  ncol = 2
)
```

However, since `partial` returns a data frame by default, using `par` would work in the following case
```{r, eval=TRUE, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
plot(partial(ozone.rf, pred.var = "Wind"), type = "l")
```

@@

### Comment 1.13
It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.

### Response 1.13
I agree, this would be a nice feature, but the problems I see are (1) not all models naturally emit a "variable importance" score; the `caret` package has come a long way in this regard, and (2) too many plots will look cluttered. At this point, I'd rather encourage the user to select which variables they would like to see plotted. I am not opposed to including this feature in the future. I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19. Thanks for the suggestion!

@@

### Comment 1.14
It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.  Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only---one can't access the data stored in `xgb.DMatrix`).

### Response 1.14
`kernlab::ksvm` does not store a copy of the training data either, as far as I can tell. As time goes on, the coverage will certainly improve, which will make it easier on the user. For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`. Also, I suspect that the package maintainers for `xgboost` (and similar packages designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model for memory reasons.

@@

### Response 1.15
This is a good suggestion and is currently under development: https://github.com/bgreenwell/pdp/blob/master/slowtests/slowtests.R. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.

# -----------Reviewer 2-----------

### Comment
The Introduction immediately introduces the PDP. You must first motivate why it is important. In a linear model, they are not important. In a linear model with simple interactions, likewise. They first become important if you have non-parametric machine-learning techniques. Further, they are not the only idea extant when trying to explain a predictor's impact on the response. This paper would benefit if you can give a bit more of a lit review here.

### Response
I expanded the introduction a bit to try and motivate why PDPs are important (the new material links directly to the Boston housing example used throughout the manuscript). I also included more background on the `plotmo` and `ICEbox` packages (and there corresponding methodologies). `plotmo` very recently introduced a `partdep` option for displaying PDPs. Because of this, more detail was added to differentiate `pdp` and motivate the use of `partial`.

I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.). In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms). I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response. Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package.

@@

## Important Clarifications Required:

### Comment
Top of p2. The `ICEbox` package allows for general construction and visualization of PDP's. However, it is limited as you only get one dimensional representations. The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.

### Response
`ICEbox` does not actually construct PDPs, rather individual curves for each observation (which can be difficults to interpret when you have millions of observations). I think flexible graphics and non-extrapoloation zones are only half of the selling point here. I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models. For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others) in parallel. I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.). Also, a new section was added to the manuscript: **User-defined prediction functions**. This section describes the new `pred.fun` argument (as suggested by a later comment) that makes `partial` even more flexible.

@@

### Comment 2.7
p3 "Error: The training..." should be wrapped.

### Response 2.7
Fixed. The text is now wrapped appropriately. Thanks for catching this!

### Comment 2.8
Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?

### Response 2.8
The sidebar (i.e., colorkey) has been labeled, but for this example only. As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for `lattice` plots. The code for doing so has been incorporated in the example in question (see pages 8-9). Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found. This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey. I don't think it would be terrible to label all the axes. However, in my option, labelling all the axes would (1) looks cluttered, and (2) require more unnecessary code that would probably distract the reader. This is the standard axis labelling scheme used in trellis plots from S-plus to R (via `lattice`).

@@

### Response 2.9
Fixed. Color key labeled "Centered logit".

### Suggestions for the software

### Comment
I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP. I realize that the default is `FALSE`. However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again. So, it is my opinion that you should take it out and force two steps: generation and illustration.

### Response
While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them first. However, I agree with your concern which is why `partial` has `plot = FALSE` as the default. Furthermore, I have made it possible to retrieve the raw PDP data even when \code{plot = TRUE}. This is because the `"trellis"` object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute. This is described in the package documentation. For example,
```{r, eval=TRUE, fig.width=5, fig.height=3}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

@@

### Comment
The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.

### Response
This is an excellent suggestion. First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.). Second, `partial` gained a new argument, `pred.fun`, where the user can provide an optional prediction function. This makes `partial` far more flexible. For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves. This is illusrated in the new section labeled **User-defined prediction functions**

### 2017-01-04 manuscript/greenwell.tex
minor edits

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{aligned}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS \\
  & \quad - 0.0064 NOX^2
\end{aligned}
\end{equation}

Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy. (We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like in Equation~\eqref{eqn:boston}. Fortunately, these models can still provide insight into the data, but it is not in the form of simple equations. For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors in the training data.

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots. \dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots. However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

@@

# Display plots side by side (Figure 3)
grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load the parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)  # register the parallel backend
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The last three lines of code simply add a label to the colorkey. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice

@@

\address{Brandon M. Greenwell\\
  Infoscitex Corporation\\
  4027 Colonel Glenn Highway\\
  Suite 210\\
  Dayton, OH 45431-1672\\
  United States of America\\}
\email{greenwell.brandon@gmail.com}

>>>>

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median home values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{aligned}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS \\
  & \quad - 0.0064 NOX^2
\end{aligned}
\end{equation}

Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy. (We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}. These models can still provide insight into the data, but it is not in the form of simple equations. For example, quantifying predictor importance has become a crucial task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots. \dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provide tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots. However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

@@

# Figure 3
grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load the parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)  # register the parallel backend
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The last three lines of code add a label to the colorkey. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice

@@

\address{Brandon M. Greenwell\\
  Infoscitex, a DCS Corporation\\
  4027 Colonel Glenn Highway\\
  Suite 210\\
  Dayton, OH 45431-1672\\
  United States of America\\}
\email{greenwell.brandon@gmail.com}
